\documentclass[aspectratio=169]{beamer}

% Theme and Fonts
\usetheme{metropolis}
\usefonttheme{professionalfonts}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{bookmark}

% Metropolis adjustments
\metroset{progressbar=none}
\setbeamerfont{frametitle}{size=\Large}
\setbeamertemplate{footline}{}

% Colors and frame title formatting
\setbeamercolor{normal text}{fg=black,bg=white}
\setbeamercolor{background canvas}{bg=white}
\setbeamercolor{frametitle}{fg=black, bg=white}

% Center frame titles
\makeatletter
\setbeamertemplate{frametitle}{
  \nointerlineskip%
  \begin{beamercolorbox}[wd=\paperwidth, sep=0.3cm, center]{frametitle}%
    \usebeamerfont{frametitle}\insertframetitle\par%
    \if\insertframesubtitle\relax%
    \else%
      \vspace{0.5em}%
      {\usebeamerfont{framesubtitle}\insertframesubtitle\par}%
    \fi%
  \end{beamercolorbox}%
}
\makeatother

% Packages
\usepackage{graphicx}
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    bookmarksnumbered=true
}
\usepackage{etoolbox}
\usepackage{media9}
\usepackage{menukeys}

\title{Building an Automated ML Pipeline — Introduction and Ethical Foundations}
\subtitle{AI Masters Capstone Project}
\author{Jonathan Agustin}
\date{November 2024}

\begin{document}

%------------------------------------------------------------
% Title Slide
%------------------------------------------------------------
%%% VOICEOVER ON
% Let's start by setting the stage. We're not just talking about building ML pipelines;
% we're talking about building them ethically, fairly, and in compliance with legal expectations.
% This presentation will show why fairness, trust, and user impact matter as much—if not more—than technical excellence.
%%% VOICEOVER OFF
\maketitle

%------------------------------------------------------------
\section{Setting the Stage}

\begin{frame}{A New Paradigm: GenAI as the Next Internet}
%%% VOICEOVER ON
% Imagine the early internet: many doubted it, called it a bubble.
% Yet it exploded in value. The same is happening with GenAI.
% The key takeaway? Trust, distribution, and seamless integration matter more than raw model sophistication.
%%% VOICEOVER OFF
\begin{itemize}
\item GenAI parallels the early internet’s transformative journey.
\item Success depends on trust, distribution, and integration, not just model strength.
\item Ethics and fairness are vital pillars for sustainable growth.
\end{itemize}
\end{frame}

\begin{frame}{Lessons from the Past Applied to GenAI}
%%% VOICEOVER ON
% Many organizations keep AI tech locked in internal experiments.
% Real impact demands delivering solutions that users trust and rely on daily.
% By embracing ethics and fairness, even non-tech companies can surpass tech giants.
%%% VOICEOVER OFF
\begin{itemize}
\item Move beyond internal pilots—user-facing solutions drive true impact.
\item Non-tech firms with trusted brands can outshine tech giants.
\item Ethics, fairness, and scale must intertwine for long-term success.
\end{itemize}
\end{frame}

%------------------------------------------------------------
\section{Ethical Imperatives and Legal Accountability}

\begin{frame}{Ethical AI Development is Not Optional}
%%% VOICEOVER ON
% AI is at an early stage, similar to the early internet. Missteps can harm millions.
% As developers, we have ethical and legal responsibilities.
% The case of United States v. Liang reminds us that "following orders" is no defense.
%%% VOICEOVER OFF
\begin{itemize}
\item AI is formative; errors at scale can harm millions.
\item Developers share ethical and legal responsibility.
\item \textbf{United States v. Liang}: Engineer held liable for unethical code despite following instructions.
\end{itemize}

\footnotesize (Ref: \href{https://www.justice.gov/opa/pr/volkswagen-engineer-pleads-guilty-his-role-conspiracy-cheat-us-emissions-tests}{DOJ Press Release})
\end{frame}

\begin{frame}{Mobley v. Workday: AI Accountability in Hiring}
%%% VOICEOVER ON
% AI vendors can be held liable for discriminatory AI outcomes.
% Mobley v. Workday: Automated decisions face the same scrutiny as human decisions.
% We must ensure fairness from the start.
%%% VOICEOVER OFF
\begin{itemize}
\item \textbf{Mobley v. Workday (2024)}: AI providers can be liable for bias in hiring algorithms.
\item Automated decisions are legally equivalent to human-made ones.
\item Technical design can create systemic discrimination if not checked.
\end{itemize}
\end{frame}

\begin{frame}{Personal Responsibility}
%%% VOICEOVER ON
% “I was just following orders” won’t protect you legally.
% You must challenge unethical directives.
% Ethical vigilance protects users, organizations, and your own career.
%%% VOICEOVER OFF
\begin{itemize}
\item “Following orders” is not a legal shield.
\item Engineers must question potentially harmful directives.
\item Uphold fairness and transparency to safeguard all stakeholders.
\end{itemize}
\end{frame}

%------------------------------------------------------------
\section{Automated ML Pipelines: Vision and Tools}

\begin{frame}{From Notebooks to Production}
%%% VOICEOVER ON
% Getting a model from a research notebook to production is hard.
% You need robust preprocessing, versioning, compliance, and more.
% Let’s build “roads” for our “race cars.”
%%% VOICEOVER OFF
\begin{itemize}
\item Transitioning from prototypes to production demands stable pipelines.
\item Preprocessing, model versioning, and compliance are crucial.
\end{itemize}
\end{frame}

\begin{frame}{Addressing Complexity and Regulation}
%%% VOICEOVER ON
% In regulated industries (finance, healthcare), compliance and auditability are critical.
% Without strong infrastructure, even stellar models fail to reach real users.
%%% VOICEOVER OFF
\begin{itemize}
\item Security, compliance, and auditability are must-haves.
\item Regulated environments add layers of complexity.
\end{itemize}
\end{frame}

\begin{frame}{Building a Comprehensive, Ethical Pipeline}
%%% VOICEOVER ON
% Automation is one stepping stone to consistent, reliable ML pipelines.
% But automation without ethics can amplify bias.
% We integrate bias checks, fairness metrics, and documentation at every stage.
%%% VOICEOVER OFF
\begin{itemize}
\item Automation: consistent preprocessing, reproducible training, smooth deployment.
\item Integrate bias detection and fairness checks, not as afterthoughts, but core components.
\end{itemize}
\end{frame}

\begin{frame}{Key Technologies}
%%% VOICEOVER ON
% Tools that help:
% - GitHub Actions for CI/CD,
% - Docker for consistent environments,
% - Terraform for scalable infrastructure,
% - Hugging Face Spaces for easy model hosting.
% Master these to build a robust ML pipeline.
%%% VOICEOVER OFF
\begin{itemize}
\item \textbf{GitHub Actions}: CI/CD for seamless integration and testing.
\item \textbf{Docker}: Consistent environments from dev to production.
\item \textbf{Terraform}: Infrastructure as code on AWS for scalability.
\item \textbf{Hugging Face Spaces}: Effortless model deployment and interfaces.
\end{itemize}
\end{frame}

%------------------------------------------------------------
\section{Real-World Compliance Examples}

\begin{frame}{FTC \& IntelliVision: Compliance Matters}
%%% VOICEOVER ON
% IntelliVision’s case: Misleading claims about facial recognition accuracy and fairness.
% A reminder that “no bias” and “high accuracy” must be backed by evidence.
%%% VOICEOVER OFF
\begin{itemize}
\item IntelliVision claimed unmatched accuracy and zero bias without proper evidence.
\item FTC scrutiny shows claims must be substantiated.
\end{itemize}
\end{frame}

%------------------------------------------------------------
\section{Operation AI Comply: The FTC’s Crackdown}

\begin{frame}{Operation AI Comply: FTC’s Enforcement Sweep}
%%% VOICEOVER ON
% The FTC’s Operation AI Comply (Sept 25, 2024): A direct assault on deceptive AI claims.
% Companies using AI hype to mislead consumers are in the crosshairs.
%%% VOICEOVER OFF
\begin{itemize}
\item FTC’s Operation AI Comply targets deceptive AI claims and schemes.
\item Multiple enforcement actions highlight that AI is no loophole for illegal practices.
\end{itemize}
\end{frame}

\begin{frame}{Key Cases Highlighted by the FTC}
%%% VOICEOVER ON
% Several cases:
% - DoNotPay claimed “AI Lawyer” services that it couldn’t deliver.
% - Ascend Ecom, Ecommerce Empire Builders, FBA Machine used AI hype for get-rich-quick schemes.
% - Rytr enabled fake reviews using AI.
% Lesson: Unsubstantiated AI claims or tools that facilitate deception invite legal action.
%%% VOICEOVER OFF
\begin{itemize}
\item \textbf{DoNotPay}: Claimed an “AI Lawyer,” settled with FTC, must warn consumers.
\item \textbf{Ascend Ecom, Ecommerce Empire Builders, FBA Machine}: Promised easy earnings via AI but delivered little.
\item \textbf{Rytr}: AI-generated fake reviews, violating consumer trust.
\end{itemize}
\end{frame}

\begin{frame}{Implications of Operation AI Comply}
%%% VOICEOVER ON
% The FTC is making it clear: using AI to mislead consumers is illegal.
% Claims must be substantiated by evidence.
% Ethical and legal compliance ensures fair competition and consumer protection.
%%% VOICEOVER OFF
\begin{itemize}
\item Deceptive AI claims = illegal.
\item Must have evidence for performance, fairness, and accuracy claims.
\item Ensures honest businesses thrive and consumers stay protected.
\end{itemize}
\end{frame}

\begin{frame}{What This Means for Developers and Businesses}
%%% VOICEOVER ON
% As developers, we must ensure transparency, evidence, and fairness.
% Document testing, training data, known biases.
% Compliance is part of the pipeline design, not a last-minute addition.
%%% VOICEOVER OFF
\begin{itemize}
\item Transparency: Document testing and limitations.
\item Accuracy: Substantiate claims, especially about bias or performance.
\item Compliance: Integrate legal awareness into the pipeline workflow.
\end{itemize}
\end{frame}

%------------------------------------------------------------
\section{Roadmap of the Series}

\begin{frame}{Our Series Roadmap}
%%% VOICEOVER ON
% This is just the start. Next, we focus on data—ethical data handling is the foundation.
% Then model training, deployment, and ensuring ongoing fairness and compliance.
%%% VOICEOVER OFF
\begin{itemize}
\item Presentation 1: Automated ML Pipeline, Ethical Foundations (Today)
\item Presentation 2: Data Preparation \& Ethical Handling
\item Presentation 3: Automating Model Training \& Fairness Checks
\item Presentation 4: Deployment Automation \& Compliance
\item Presentation 5: Code Quality, Security \& Ethical Governance
\end{itemize}
\end{frame}

%------------------------------------------------------------
\section{Next Steps}

\begin{frame}{Next Steps}
%%% VOICEOVER ON
% Up next: Laying a strong, ethical foundation at the data layer.
% Without trustworthy, representative data, no pipeline can be truly fair or compliant.
% We’ll learn how to ensure data quality, fairness, and transparency from the ground up.
%%% VOICEOVER OFF
\begin{itemize}
\item Next: Ethical data preparation—ensuring trustworthy, representative datasets.
\item Building pipelines that are technically sound, ethically robust, and legally compliant.
\end{itemize}
\end{frame}

\begin{frame}{Additional Resources and References}
%%% VOICEOVER ON
% For deeper understanding, consider reviewing FTC materials and legal frameworks.
% Follow regulatory updates, as the landscape evolves quickly.
% Stay informed to stay compliant.
%%% VOICEOVER OFF
\begin{itemize}
\item FTC Press Releases and Notices: Learn from real enforcement actions.
\item Explore \href{https://www.ftc.gov/}{FTC.gov} for guidelines and updates.
\item Operation AI Comply: Critical lessons on accountability.
\end{itemize}
\end{frame}

\begin{frame}{Closing Thoughts}
%%% VOICEOVER ON
% Ethical AI is about responsibility and foresight.
% By integrating fairness, transparency, and compliance, we protect ourselves, our users, and our organizations.
% This journey ensures we build AI pipelines that truly serve humanity.
%%% VOICEOVER OFF
\begin{itemize}
\item Ethical AI isn’t optional—it’s essential.
\item Responsible innovation safeguards users, developers, and businesses.
\item Let’s build pipelines that uphold these values at every step.
\end{itemize}
\end{frame}

\end{document}
